---
title: "EDA of Alcohol Consumption among Students"
author: Ilia Ozhmegov, Olena Horyn
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document:
    df_print: paged
    highlight: pygments
    fontsize: 11pt
    number_section: true
    toc_depth: 2
    toc_float: true
    toc: yes
    fig_caption: true
    theme: cosmo
  pdf_document:
    latex_engine: xelatex
    df_print: kable
    highlight: pygments
    number_section: true
    toc: yes
    fig_caption: true
---

```{r Setting template options, include=FALSE}

# knitr::opts_chunk$set(message=TRUE, warning = TRUE, echo = TRUE)
knitr::opts_chunk$set(message=FALSE, warning = FALSE, echo = TRUE)
knitr::opts_chunk$set(fig.show = "hold", fig.align = "center")

```


```{r Loading libraries, include=FALSE}

# library(aider)
library(tidyverse)
setwd('..')
source("load_data.R")

matlab = tibble(blue='#1F77B4', orange='#FF7F0E')

```
# Introduction 

Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text.  Introduction text. 

# Data Preparation

It is done in `load_data.R` file.

# Correlation Bar Plot

Text.

```{r}
  int_df <- lapply(full_df, as.integer)
  n = length(int_df)
  
  c = rep(0, n)
  for(i in 1:n) c[i] = cor(int_df$G3, int_df[[i]]) %>% abs()

  cor_df = data.frame(correlation=c, name=colnames(full_df)) 
  
  cor_df %>% 
    filter(!str_detect(name, "^G")) %>% 
    ggplot(aes(x=correlation, y=reorder(name, correlation))) + 
    geom_bar(stat="identity", aes(fill=str_detect(name, "alc$"))) +
    scale_fill_manual(values = c(matlab$blue, matlab$orange) ) +
    theme_bw() +
    theme(legend.position="none") +
    labs(y = "Variable", x = "Correlation", title = "Final Grade Correlation")
```

# A closer look at alcohol consumption

Здесь могла быть ваша реклама. Здесь могла быть ваша реклама. Здесь могла быть ваша реклама. 

```{r}

full_df %>% 
  ggplot(aes(x=Salc, y=G3, fill=Salc)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position="none") +
  # scale_fill_manual(values=grades_colours) +
  xlab("Alcohol Consumption") +
  ylab("Grade") +
  ggtitle("Final period grade")

```




```{r}

full_df %>% 
  ggplot(aes(x= Salc, y=G3, fill=Salc)) +
  geom_violin() +
  theme_bw() +
  theme(legend.position="none") +
  # scale_fill_manual(values=grades_colours)
  labs(y = "Final Grade", x = "Alcohol consumption", title = "Final period grade")
```


```{r}
  full_df %>% 
    ggplot(aes(x=Salc, G3_d)) +
    geom_jitter(aes(colour = Salc)) + 
    theme_bw() +
    theme(legend.position="none") +
    geom_hline(yintercept='C',      color = "black") + 
    geom_vline(xintercept='Medium', color = "black") +
    labs(y = "Final Grade (Descrete)", x = "Alcohol consumption", title = "Final period grade")
```

Here clearly seen that there is extreamly poor dependece between alcohol 
consumption and decent grades. However, the conditional probability to have 
decent grade is little bit higher when alcohol consumption is not `High$`.